Vagueness, Noise, and Signaling

### 2015-09-30 paper/version_04/paper.tex
section 3 edits

@@

Fix a signaling game with finite states $\States$ and messages $\Messgs$. We consider an
interpretation game in which the receiver chooses states in $\States$ in response to
messages. Let $\Pr(\cdot) \in \Delta(\States)$ be the prior distribution over states and
$\Util \mycolon \States \times \Messgs \times \States \rightarrow \mathds{R}$ the utility
function shared by senders and receivers in the population. A \emph{behavioral strategy} is a
function that maps an agent's choice points to a probability distribution over available
choices.\footnote{Our focus is on behavioral strategies, not \emph{mixed strategies}, i.e.,
  probability distributions over functions from each choice point to an act, such as
  $s \in \Delta(\Messgs^{\States})$.  Dynamics on behavioral strategies assume that agents can
  adapt their behavior locally, i.e., independently at each choice point. Our
  focus on behavioral strategies greatly reduces the complexity of the dynamic and simplifies
  numerical simulations. But it also seems the more plausible choice for imitation-based update
  protocols: agents only observe how, on some occasion, some other agent behaved in \emph{one}
  particular situation, not how that agent would behave in \emph{all} relevant choice
  situations; they imitate the use of a single word, so to speak, not a whole lexicon. (See
  \citet{Cressman2003:Evolutionary-Dy} for more on the difference between dynamics on mixed or
  behavioral strategies.)} The sender's behavioral strategies are functions
$\Sstrat \in \Delta(\Messgs)^\States$, thus mapping each state $\state \in \States$ to a
probability of each message $\messg \in \Messgs$ being sent in $\state$; the receiver's are
functions $\Rstrat \in \Delta(\States)^\Messgs$, thus mapping each message $\messg \in \Messgs$
to a probability of each interpretation $\state \in \States$ being chosen in response to
$\messg$. Although behavioral strategies are probabilistic, evolutionary modeling usually
imagines that every individual agent has a non-probabilistic strategy. Behavioral strategies
then capture average population behavior. In a virtually infinite population, the number
$\Sstrat(\messg \probbar \state)$, for instance, is then the probability that a randomly
sampled sender would send message $\messg$ if the actual state was $\state$. Similarly,
$\Rstrat(\state \probbar \messg)$ is then the probability with which a randomly sampled
receiver interprets $\messg$ as $\state$.

The \emph{expected utility} of choices at each choice point is:\footnote{The notation
  $\propto$, for ``proportional to'', that is used here and hereafter means that the right-hand
  side must possibly yet be normalized so as to have probabilities sum to one. In general,
  writing $P(x) \propto f(x)$ for any function $f \mycolon X \rightarrow \mathds{R}$ is
  shorthand for $P(x) = \frac{f(x)}{\sum_{x' \in X} f(x')}$. }
\begin{align*}
  \EU(\messg, \state, \Rstrat) & = \sum_{\state \in \States}
  \Rstrat(\state' \probbar \messg) \myts U(\state, \messg, \state') \\
  \EU(\state', \messg, \Sstrat) & = \sum_{\state \in
    \States} P(\state \probbar \messg)  \myts
  U(\state, \messg, \state')\,, \text{\ \  where } &
P(\state \probbar \messg) & \propto \Pr(\state) \myts \Sstrat(\messg \probbar \state) \,.
\end{align*}
Expected utilities are sums, over all possible concrete outcomes, of the utilities of these
outcomes, weighted by how likely they are to occur. For instance, the expected utility for the
sender of choosing message $\messg$, given tate $\state$ and a receiver strategy $\Rstrat$ is
the sum, for each act $\act$, of the probability that some random receiver will choose
$\act$ given $\messg$ times the actual utility of triple $\state$, $\messg$ and $\act$.

The discrete-time replicator dynamic tracks changes in frequency of choices in the population
as proportional to their expected utilities
\citep[e.g.][]{HofbauerSigmund1998:Evolutionary-Ga}:\footnote{The formulation given in
  Equation~(\ref{eq:RDDiscrete}) is adequate only for cases like the one we will be looking at
  in Section~\ref{sec:exploring-rdd}, where utilities are always non-negative and expected
  utilities are always positive.}
\begin{align}
  \label{eq:RDDiscrete}
  \Sstrat'(\messg \probbar \state) & \propto \Sstrat(\messg \probbar \state) \cdot
    \EU(\messg, \state,\Rstrat)\,, & \Rstrat'(\act \probbar \messg) & \propto \Rstrat(\act \probbar \messg) \cdot
    \EU(\act, \messg,\Sstrat)\,.   
\end{align}
For a given choice point, say a state $\state$, the probability of seeing $\messg$ played by an
average agent in the population \emph{after} the update is proportional to the probability of
seeing it \emph{before} the update, which is $\Sstrat(\messg \probbar \state)$, times the
expected utility of $\messg$ at state $\state$. Intuitively put, frequencies of choices change
by a gradient of current frequency and a measure of how good they are.

\subsection{A noisy replicator dynamic}
\label{sec:noisy-repl-dynam}

@@

Probabilistic imitation of success presupposes that agents make no mistakes when observing
states. This may not always be an appropriate assumption. Especially for sim-max games, where,
intuitively, some states can be fairly similar and therefore likely to be confused for one
another. Appendix~XYZ shows how to derive a generalization of the replicator dynamic from the
assumption that probabilistic imitation of success is affected by two sources of probabilistic
confusion of states:

@@

The mean changes in the population frequencies of choices under noise-affected imitation of
success can be covered by the following discrete-time formulation:

@@

This looks very much like the discrete-time formulation of the standard replicator dynamic in
(\ref{eq:RDDiscrete}), but there are differences (the Appendix~XYZ for full definitions and
formal detail). Firstly, expected utility here takes stochastic confusability of states into
account in a natural way. Secondly, where the standard replicator dynamic had probabilities
$\Sstrat(\messg \probbar \state)$ and $\Rstrat(\state \probbar \messg)$, we now have
$P_o(\messg \probbar \state)$ and $P_o(\state \probbar \messg)$ respectively. This is the major
difference, so we should dwell on it for a brief moment.

Look at the sender case first, where $\Sstrat(\messg \probbar \state)$ is the probability with
which $\messg$ is played in response to actual state $\state$ on average in the population. If
agents may not perceive the actual state, this should also affect what they observe when trying
to imitate another agents behavior. $P_o(\messg \probbar \state)$ is therefore the probability,
given that a potential imitator has observed $\state$, that he would observe a random agent
play $\messg$, which he may or may not choose to imitate. Similarly for the receiver, we are
not looking at occurrence probability $\Rstrat(\state \probbar \messg)$ but at the probability
$P_o(\state \probbar \messg)$ with which an agent who observes $\messg$ observes another agent
choose interpretation $\state$.

Obviously, $P_o(\messg \probbar \state)$ and $P_o(\state \probbar \messg)$ are to be derived
from, among others, probabilistic confusion of states $P_o(\mystate{o} \probbar \mystate{a})$
and $P_r(\mystate{r} \probbar \mystate{o})$. Details are given in Appendix~XYZ. If there is no
noise, $P_o(\messg \probbar \state)$ reduces to $\Sstrat(\messg \probbar \state)$ and
$P_o(\state \probbar \messg)$ reduces to $\Rstrat(\state \probbar \messg)$. But if there is,
imitation probabilities must take confusability of states into account. In this sense, the
noisy replicator dynamic in (\ref{eq:nRD}) conservatively extends the classic case in
(\ref{eq:RDDiscrete}).

>>>>

\section{Imitation with confusability of states}
\label{sec:repl-diff-dynam}

@@

Fix a signaling game with finite states $\States$ and messages $\Messgs$. We consider an
interpretation game in which the receiver chooses states in $\States$ in response to
messages. Let $\Pr(\cdot) \in \Delta(\States)$ be the prior distribution over states and
$\Util \mycolon \States \times \Messgs \times \States \rightarrow \mathds{R}$ the utility
function shared by senders and receivers in the population. A \emph{behavioral strategy} is a
function that maps an agent's choice points to a probability distribution over available
choices.\footnote{Our focus is on behavioral strategies, not \emph{mixed strategies}, i.e.,
  probability distributions over functions from each choice point to an act, such as
  $s \in \Delta(\Messgs^{\States})$.  Dynamics on behavioral strategies assume that agents can
  adapt their behavior locally, i.e., independently at each choice point. Our focus on
  behavioral strategies greatly reduces the complexity of the dynamic and simplifies numerical
  simulations. But it also seems the more plausible choice for imitation-based update protocols
  of the kind we consider here: agents only observe how, on some occasion, some other agent
  behaved in \emph{one} particular situation, not how that agent would behave in \emph{all}
  relevant choice situations; they imitate the use of a single word, so to speak, not a whole
  lexicon. (See \citet{Cressman2003:Evolutionary-Dy} for more on the difference between
  dynamics on mixed or behavioral strategies.)} The sender's behavioral strategies are
functions $\Sstrat \in \Delta(\Messgs)^\States$, thus mapping each state $\state \in \States$
to a probability of each message $\messg \in \Messgs$ being sent in $\state$; the receiver's
are functions $\Rstrat \in \Delta(\States)^\Messgs$, thus mapping each message
$\messg \in \Messgs$ to a probability of each interpretation $\state \in \States$ being chosen
in response to $\messg$. Although behavioral strategies are probabilistic, evolutionary
modeling usually imagines that every individual agent has a non-probabilistic
strategy. Behavioral strategies then capture average population behavior. Assuming a virtually
infinite population, the number $\Sstrat(\messg \probbar \state)$, for instance, is then the
probability that a randomly sampled sender would send message $\messg$ if the actual state was
$\state$. Similarly, $\Rstrat(\state \probbar \messg)$ is then the probability with which a
randomly sampled receiver interprets $\messg$ as $\state$.

The \emph{expected utility} of choices at each choice point is:\footnote{The notation
  $\propto$, for ``proportional to'', that is used here and hereafter means that the right-hand
  side must possibly yet be normalized so as to have probabilities sum to one. In general,
  writing $P(x) \propto f(x)$ for any function $f \mycolon X \rightarrow \mathds{R}$ is
  shorthand for $P(x) = \frac{f(x)}{\sum_{x' \in X} f(x')}$. }
\begin{align*}
  \EU(\messg, \state, \Rstrat) & = \sum_{\state \in \States}
  \Rstrat(\state' \probbar \messg) \myts U(\state, \messg, \state') \\
  \EU(\state', \messg, \Sstrat) & = \sum_{\state \in
    \States} P(\state \probbar \messg)  \myts
  U(\state, \messg, \state')\,, \text{\ \  where } &
P(\state \probbar \messg) & \propto \Pr(\state) \myts \Sstrat(\messg \probbar \state) \,.
\end{align*}
Expected utilities are sums, over all possible concrete outcomes, of the utilities of these
outcomes, weighted by how likely they occur. For instance, the expected utility for the sender
of choosing message $\messg$, given state $\state$ and a receiver strategy $\Rstrat$ is the
sum, for each interpretation $\state'$, of the probability that some random receiver will
choose $\state'$ given $\messg$ times the actual utility of triple $\state$, $\messg$ and
$\state'$.

The discrete-time replicator dynamic tracks changes in frequency of choices in the population
as proportional to their expected utilities
\citep[e.g.][]{HofbauerSigmund1998:Evolutionary-Ga}:\footnote{The formulation given in
  Equation~(\ref{eq:RDDiscrete}) is adequate only for cases like the one we will be looking at
  in Section~\ref{sec:exploring-rdd}, where utilities are always non-negative and expected
  utilities are always positive.}
\begin{align}
  \label{eq:RDDiscrete}
  \Sstrat'(\messg \probbar \state) & \propto \Sstrat(\messg \probbar \state) \myts
    \EU(\messg, \state,\Rstrat)\,, & \Rstrat'(\act \probbar \messg) & \propto \Rstrat(\act \probbar \messg) \myts
    \EU(\act, \messg,\Sstrat)\,.   
\end{align}
For a given choice point, say a state $\state$, the probability of seeing $\messg$ played by an
average agent in the population \emph{after} the update is proportional to the probability of
seeing it \emph{before} the update, which is $\Sstrat(\messg \probbar \state)$, times the
expected utility of $\messg$ at state $\state$. Intuitively put, frequencies of choices change
by a gradient of current frequency and a measure of how good they are.

\subsection{Noisy replicator dynamic}
\label{sec:noisy-repl-dynam}

@@

Probabilistic imitation of success presupposes that agents make no mistakes when observing
states. This may not always be an appropriate assumption, especially for sim-max games, where,
some states can be fairly similar and therefore likely to be confused for one
another. Appendix~XYZ shows how to derive a generalization of the replicator dynamic from the
assumption that imitation of success is affected by two sources of probabilistic noise:

@@

\noindent Noise-free imitation of success is the special case where $P_o(\mystate{o} \probbar
\mystate{a}) = 1$ iff $\mystate{a} = \mystate{o}$ and $P_r(\mystate{r} \probbar
\mystate{i}) = 1$ iff $\mystate{i} = \mystate{r}$. In that case, the population-level impact of
imitation is described by the replicator equation in (\ref{eq:RDDiscrete}).

If noise is possibly non-trivial, the mean changes in population frequencies of choices
under noise-affected imitation of success can be covered by the following discrete-time
formulation:

@@

\noindent This looks very much like the discrete-time formulation of the standard replicator
dynamic in (\ref{eq:RDDiscrete}), but there are differences (the Appendix~XYZ for full
definitions and formal detail). Firstly, expected utility here takes stochastic confusability
of states into account in a natural way. Secondly, where the standard replicator dynamic had
probabilities $\Sstrat(\messg \probbar \state)$ and $\Rstrat(\state \probbar \messg)$, we now
have $P_o(\messg \probbar \state)$ and $P_o(\state \probbar \messg)$ respectively. This is the
major difference, so we should dwell on it for a brief moment.

Look at the sender case first. Is some agent $X$ gets a chance to revise his behavior, he
observes a state $\state$ and, at the same time, that some agent $Y$ chose $\messg$. But if $X$
perceives $\state$ that does not necessarily mean that the actual state is $\state$, and
similarly, that agent $Y$ also perceived state $\state$. Hence, where the replicator dynamic in
(\ref{eq:RDDiscrete}), derived from noise-free imitation, has the probability
$\Sstrat(\messg \probbar \state)$ with which $\messg$ is played in response to actual state
$\state$ on average in the population, we substitute $P_o(\messg \probbar \state)$: the
probability, given that a potential imitator has observed $\state$, that he would observe a
random agent play $\messg$. Similar considerations apply to the receiver case: we are not
looking at occurrence probability $\Rstrat(\state \probbar \messg)$ but at the probability
$P_o(\state \probbar \messg)$ with which an agent who observes $\messg$ observes another agent
choose interpretation $\state$.

Obviously, $P_o(\messg \probbar \state)$ and $P_o(\state \probbar \messg)$ are to be derived
from, among others, the assumed probabilistic confusion rates
$P_o(\mystate{o} \probbar \mystate{a})$ and $P_r(\mystate{r} \probbar \mystate{o})$. Details
are given in Appendix~XYZ. If there is no noise, $P_o(\messg \probbar \state)$ reduces to
$\Sstrat(\messg \probbar \state)$ and $P_o(\state \probbar \messg)$ reduces to
$\Rstrat(\state \probbar \messg)$. But if there is, we must take
confusability of states into account. In this sense, the noisy replicator dynamic in
(\ref{eq:nRD}) conservatively extends the classic case in (\ref{eq:RDDiscrete}).
